\documentclass[10pt]{article}
\pagestyle{empty}               % no page numbers, thanks.
\usepackage{hyperref}
\usepackage{ltablex}
\usepackage{ifthen}
\usepackage[tiny]{titlesec}
\usepackage[backend=biber,style=numeric-comp,sorting=none,giveninits=true,maxnames=20,eprint=false,isbn=false]{biblatex}
\usepackage{xstring}

% Add bib file to make bibliography information
\addbibresource{../../Scholarship/scholarship.bib}


% Set page size
\textwidth = 6.5 in
\textheight = 9 in
\oddsidemargin = 0.0 in
\evensidemargin = 0.0 in
\topmargin = 0.0 in
\headheight = 0.0 in
\headsep = 0.0 in
\parindent= 0 in  

\title{Faculty Activity Report}
\author{}

\begin{document}
\maketitle

\section{INSTRUCTIONS}
\begin{itemize}
\item THE DATA BELOW IS PROVIDED AS A CONVENIENCE.  IF IT IS INCORRECT OR MISSING ITEMS, YOU ARE RESPONSIBLE FOR UPDATING IT.
\item The data is taken either from PeopleSoft or from other University or publicly available sources such as ORCID or Google Scholar or cross-ref.
\item For the automatic retrieval of scholarly works, you need to have have claimed your Google Scholar Profile and/or your ORCID profile.  This is highly recommended as it brings your research greater visibility.  Instructions for claiming your profile are here:
\begin{itemize}
\item Google: \url{https://scholar.google.com/intl/en/scholar/citations.html}
\item ORCID: \url{https://support.orcid.org/hc/en-us/articles/360006897454-How-do-I-register-for-an-ORCID-ID}
\end{itemize}
\item If after receiving this you claim your profiles, send an e-mail to far@clarkson.edu and your FAR will be regenerated with automatically imported data.
\item If you  maintain own bibliographic database, you can output a ``.bib'' file, and send it to far@clarkson.edu and it will be used to generate the lists of scholarly publications.
\item The division of journal papers, conference papers, etc...  by Google Scholar and ORCID  is not very reliable.  You may have to move publications into different categories
\item Another note about Google.  Google blocks requests to download full bibliographic data, so the biliographic record has to be completed with other sources.  For this reason some Google Scholar items may not get imported.
\item Many sections below are not relevant to all faculty.  Leave sections blank that are not relevant or you feel are unimportant to report. 
\item The data in the "SERVICE" section was gathered from many University sources and were in a wide range of formats so inconsistencies are likely.
\item If ORCID has reviewing data, it is provided in the section "SERVICE - Professional".  This is provided for your information only.  Chairs are aware that many journals do not currently report to ORCID.
\item M.S. theses were not being cataloged in ProQuest until very recently (only Ph.D. theses).  Student MS theses can be entered by hand in the section ``GRADUATE ADVISEES''
\item The section "PROPOSALS" come from PeopleSoft.  Proposals appear if the submission date recorded by SRS was in this calendar year.
\item The section "GRANTS" also come from PeopleSoft.  Grants appear if the active period of the grant overlaps this calendar year.  The dollar amounts are the amount received to date.
\item The section ``NARRATIVE''  is for you to provide context to the data and to describe activities that don't fall into the traditional categories.
\item The section ``GOALS'' is for you to describe your goals for the upcoming calendar year.
\item If something seems totally out of whack, send a note to far@clarkson.edu explaining the problem with as much detail as possible.

\end{itemize}

\newpage
\section{SCHOLARLY OUTPUT}
\subsection{Journal Papers}
\input{Tables_far/Journal.tex}

\subsection{arXiv Papers}
\input{Tables_far/arXiv.tex}

\subsection{Refereed Conference Papers}
\input{Tables_far/Refereed.tex}

\subsection{Books \& Book Chapters}
\input{Tables_far/Book.tex}

\subsection{Conferences}
\input{Tables_far/Conference.tex}

\subsection{Patents}
\input{Tables_far/Patent.tex}

\subsection{Invited Talks}
\input{Tables_far/Invited.tex}

\subsection{Other Scholarly Works}

\section{AWARDS}
\subsection{Personal Awards}
\input{Tables_far/PersonalAwards.tex}

\subsection{Student Awards}
\input{Tables_far/StudentAwards.tex}

\section{SERVICE}
\subsection{University}
\input{Tables_far/Service.tex}
\input{Tables_far/ProspectiveVisits.tex}

\subsection{Professional}
\input{Tables_far/Reviews.tex}

\subsection{Community}

\section{UNDERGRADUATE ADVISING}
\input{Tables_far/AdviseeCounts.tex}
\input{Tables_far/AdvisingEvals.tex}

\section{GRADUATE ADVISEES \& THESES}
\input{Tables_far/GradAdvisees.tex}

\section{UNDERGRADUATE RESEARCH}
\input{Tables_far/UndergradResearch.tex}

\section{TEACHING}
\subsection{Teaching Evaluations}
This table includes all classes taught that had reported teaching evaluations.  Question 19 is ``Overall, how would you rate this instructor?''  Question 20 is ``Overall, how would you rate this course?''. The scores are out of 5 and the University average evaluation is approximately 4.1. 
\input{Tables_far/Teaching.tex}

\subsection{Course Development}

\subsection{Innovative Activities}

\section{CLINICAL WORK}
Please provide: job title, workplace info (name, address, supervisor), client populations (including age groups and conditions), average hours per week, and a general description of work.

\section{CONSULTING ACTIVITIES}
Please provide: job title, workplace info (company name, address, supervisor) average hours per week, and a general description of work.

\section{OUTREACH ACTIVITIES}

\section{PROFESSIONAL DEVELOPMENT ACTIVITIES}
\subsection{Licensure and Advanced Certification}
\subsection{Continuing Education}
\subsection{Teaching-Related Professional Development}

\section{ACTIVE GRANTS}
\input{Tables_far/Expenditures.tex}
\input{Tables_far/Grants.tex}

\section{PROPOSALS SUBMITTED}
\input{Tables_far/Proposals.tex}

\section{NARRATIVE}

\section{GOALS}

\end{document}

